**The status of the microbial census: an update**

### 2016-01-20 doc/notebook/Archaeal_data_acquisition.Rmd
Put all OTU and method information in metadata files. Will do rarefaction externally

write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/archaea.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
```

@@

write.table(archaea_emirge_metadata, file="data/mothur/archaea.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
```

@@

```{r}
#here we'll read in the list file and extract the OTU assignments for the 0.03
#cutoff
list_text <- scan(file="data/mothur/all_archaea.filter.unique.precluster.an.list", sep="\n", what="")[5]

#let's remove the otu label and the number of OTUs at this cutoff
list_otus <- unlist(strsplit(list_text, "\t"))[-c(1,2)]
n_otus <- length(list_otus)

@@

#now we get the OTU number for each sequence in the dataset
otu_assignment <- unlist(lapply(1:n_otus, split_otu))

@@

#let's get the metadata file and paste in the OTU and taxonomy data
metadata <- read.table(file="data/mothur/archaea.good.metadata", header=T, stringsAsFactors=FALSE)
metadata$otu <- otu_assignment[as.character(rownames(metadata))]
metadata$taxonomy <- taxonomy_data[as.character(rownames(metadata))]

@@

#... and we output the metadata file...
write.table(metadata, file="data/process/archaea.v123.metadata", sep="\t")

#... and we output a group file to use for splitting up the list file into the individual categories...
write.table(file="data/mothur/archaea.all_category.groups", x=cbind(rownames(metadata), metadata$category), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")

#... and we pool the categories by environment
env_categories <- rep(NA, nrow(metadata))
env_categories[metadata$category == "AE"] <- "aerosol"
env_categories[metadata$category == "OT"] <- "other"
env_categories[metadata$category %in% c("AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "AQS")] <- "aquatic"
env_categories[metadata$category %in% c("BD", "BF", "BI", "BO", "BP")] <- "built"
env_categories[metadata$category %in% c("PO", "PR", "PS")] <- "plant"
env_categories[metadata$category %in% c("SA", "SD", "SO", "SP")] <- "soil"
env_categories[metadata$category %in% c("ZA", "ZN", "ZO", "ZV")] <- "zoonotic"
write.table(file="data/mothur/archaea.env_category.groups", x=cbind(rownames(metadata), env_categories), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
```

One thing we'd like to look at are the rarefaction curves for each set of group files and the entire dataset. To do this we'll run the mothur command like so..

```{r, engine="bash"}
mothur "#set.dir(input=data/mothur, output=data/mothur);
        make.shared(list=all_archaea.filter.unique.precluster.an.list, group=archaea.all_category.groups, label=unique-0.03-0.05-0.10-0.20);
        system(mv data/mothur/all_archaea.filter.unique.precluster.an.shared data/mothur/all_archaea.all_categories.shared);
        rarefaction.single(shared=all_archaea.all_categories.shared, outputdir=data/process)"

mothur "#set.dir(input=data/mothur, output=data/mothur);
        make.shared(list=all_archaea.filter.unique.precluster.an.list, group=archaea.env_category.groups, label=unique-0.03-0.05-0.10-0.20);
        system(mv data/mothur/all_archaea.filter.unique.precluster.an.shared data/mothur/all_archaea.env_category.shared);
        rarefaction.single(shared=all_archaea.env_category.shared, outputdir=data/process)"

mothur "#rarefaction.single(list=data/mothur/all_archaea.filter.unique.precluster.an.list, label=unique-0.03-0.05-0.10-0.20, outputdir=data/process);"

rm data/mothur/*rabund data/mothur/mothur*logfile mothur*logfile
```

>>>>

spread_sheet_good$date <- "2013-07-25"

write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/archaea.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)

#we'll output the sequence names so the sequences aren't included in the
#rarefaction analysis
write(rownames(spread_sheet_good), file="data/mothur/archaea.sc_emirge.accnos")
```

@@

write.table(archaea_emirge_metadata, file="data/mothur/archaea.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)

#we'll output the sequence names so the sequences aren't included in the
#rarefaction analysis
write(rownames(archaea_emirge_metadata), file="data/mothur/archaea.sc_emirge.accnos", append=T)
```

@@

```{r}
#here we'll read in the list file and extract the OTU assignments for the
# unique, 0.03, 0.05, 0.10, 0.20 cutoffs
list_text <- scan(file="data/mothur/all_archaea.filter.unique.precluster.an.list", sep="\n", what="")[c(2,5,7,12,22)]
names(list_text) <- c("0.00", "0.03", "0.05", "0.10", "0.20")

#let's remove the otu label and the number of OTUs at each cutoff
parseList <- function(line){

	list_otus <- unlist(strsplit(line, "\t"))[-c(1,2)]
	n_otus <- length(list_otus)

@@

	#now we get the OTU number for each sequence in the dataset
	unlist(lapply(1:n_otus, split_otu))
}
otu_data <- sapply(list_text, parseList)

@@

#let's get the metadata file and paste in the OTU and taxonomy data
metadata <- read.table(file="data/mothur/archaea.good.metadata", header=T, stringsAsFactors=FALSE)
metadata <- cbind(metadata, otu_data[as.character(rownames(metadata)),])
metadata$taxonomy <- taxonomy_data[as.character(rownames(metadata))]

@@

#... and we pool the categories by environment
metadata$environment <- rep(NA, nrow(metadata))
metadata$environment[metadata$category == "AE"] <- "aerosol"
metadata$environment[metadata$category == "OT"] <- "other"
metadata$environment[metadata$category %in% c("AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "AQS")] <- "aquatic"
metadata$environment[metadata$category %in% c("BD", "BF", "BI", "BO", "BP")] <- "built"
metadata$environment[metadata$category %in% c("PO", "PR", "PS")] <- "plant"
metadata$environment[metadata$category %in% c("SA", "SD", "SO", "SP")] <- "soil"
metadata$environment[metadata$category %in% c("ZA", "ZN", "ZO", "ZV")] <- "zoonotic"

#let's flag each sequence to indicate how each was generated
source("code/partition_data.R")
metadata$pcr <- is_pcr(metadata)
metadata$cultured <- is_cultured(metadata)
metadata$single_cell <- is_single_cell(metadata)
metadata$emirge_pcr <- is_emirge_pcr(metadata)
metadata$emirge_metag <- is_emirge_metag(metadata)

#... and we output the metadata file...
write.table(metadata, file="data/process/archaea.v123.metadata", sep="\t")
```

Viola! We'll upload data/process/archaea.v123.metadata to FigShare and start analyzing the data for the paper.

### 2016-01-20 doc/notebook/Bacterial_data_acquisition.Rmd
Put all OTU and method information in metadata files. Will do rarefaction externally

Notes...

* Looks like an archaeal sequence slipped through
* Will want to get rid of sequences with large number of Ns in them - how large?

@@

write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/bacteria.good.metadata", quote=TRUE, na="NA", sep="\t", col.names=F, append=T)
```

@@

write.table(bacteria_emirge_metadata, file="data/mothur/bacteria.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)
```

@@

```{r}
#here we'll read in the list file and extract the OTU assignments for the 0.03
#cutoff (for some reason mothur was repeating the header line for each OTU threshold)
list_text <- scan(file="data/mothur/all_bacteria.filter.unique.precluster.an.list", sep="\n", what="")

list_text_03 <- list_text[grepl("^0\\.03", list_text)]

#let's remove the otu label and the number of OTUs at this cutoff
list_otus <- unlist(strsplit(list_text_03, "\t"))[-c(1,2)]
n_otus <- length(list_otus)

@@

#let's get the metadata file and paste in the OTU and taxonomy data
metadata <- read.table(file="data/mothur/bacteria.good.metadata", header=T, stringsAsFactors=FALSE)
metadata$otu <- otu_assignment[as.character(rownames(metadata))]
metadata$taxonomy <- taxonomy_data[as.character(rownames(metadata))]

@@

#correct typos:
metadata$category[metadata$category == 'AF'] <- 'AE'
metadata$category[metadata$category == 'AQBF'] <- 'AQB'
metadata$category[metadata$category == 'AQD'] <- 'AQO'
metadata$category[metadata$category == 'AQHS'] <- 'AQH'
metadata$category[metadata$category == 'AQS'] <- 'AQMS'
metadata$category[metadata$category == 'AWF'] <- 'AQF'
metadata$category[metadata$category == 'DS'] <- 'BS'
metadata$category[metadata$category == 'S'] <- 'SO'
metadata$category[metadata$category == 'SI'] <- 'BI'
metadata$category[metadata$category == 'V'] <- 'BO'
metadata$category[metadata$category == 'VD'] <- 'BD'
metadata$category[metadata$category == 'ZOZO'] <- 'ZO'
metadata$category[metadata$category == 'AH'] <- 'AQH'
metadata$category[metadata$category == 'AQOS'] <- 'AQS'
metadata$category[metadata$category == 'BS'] <- 'BD'
metadata$category[metadata$category == 'AQS'] <- 'AQFS'
metadata$category[metadata$category == 'ZL'] <- 'ZV'
metadata$category[metadata$category == 'SR'] <- 'PR'
metadata$category[metadata$category == 'M'] <- 'AQM'
metadata$category[metadata$category == 'AQFM'] <- 'AQFS'
metadata$category[metadata$category == 'SV'] <- 'ZV'
metadata$category[metadata$category == 'ZI'] <- 'ZA'

table(metadata$category)

#    AE    AQB   AQBS    AQF   AQFS    AQH    AQI    AQM   AQMS    AQO     BD
#  3473   1094    390  59220  23979  10861   2073 134788  27811   8765  33152
#    BF     BI     BO     BP     OT     PO     PR     PS     SA     SD     SO
# 11813  16598   8556  38710  19417   9753  19695   4892  10051   3059  60335
#    SP     ZA     ZN     ZO     ZV
#  1922  13209   7476  10855 794492

# need to change the name of this sequence to make it clear that it was not cultured
# rownames(metadata)[rownames(metadata) == 'AJ224540.MxeCultu'] <- 'AJ224540.UncMxeCultu'
# doing this here screws up the list file - need to do this in the actual
# analysis

#... and we output the metadata file...
write.table(metadata, file="data/process/bacteria.v123.metadata", sep="\t")

#... and we output a group file to use for splitting up the list file into the individual categories...
write.table(file="data/mothur/bacteria.all_category.groups", x=cbind(rownames(metadata), metadata$category), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")

#... and we pool the categories by environment
env_categories <- rep(NA, nrow(metadata))
env_categories[metadata$category == "AE"] <- "aerosol"
env_categories[metadata$category == "OT"] <- "other"
env_categories[metadata$category %in% c("AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "AQS")] <- "aquatic"
env_categories[metadata$category %in% c("BD", "BF", "BI", "BO", "BP")] <- "built"
env_categories[metadata$category %in% c("PO", "PR", "PS")] <- "plant"
env_categories[metadata$category %in% c("SA", "SD", "SO", "SP")] <- "soil"
env_categories[metadata$category %in% c("ZA", "ZN", "ZO", "ZV")] <- "zoonotic"
write.table(file="data/mothur/bacteria.env_category.groups", x=cbind(rownames(metadata), env_categories), quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
```

One thing we'd like to look at are the rarefaction curves for each set of group files and the entire dataset. To do this we'll run the mothur command like so..

```{r, engine="bash"}
mothur "#set.dir(input=data/mothur, output=data/mothur);
        make.shared(list=all_bacteria.filter.unique.precluster.an.list, group=bacteria.all_category.groups, label=unique-0.03-0.05-0.10-0.20);
        system(mv data/mothur/all_bacteria.filter.unique.precluster.an.shared data/mothur/all_bacteria.all_categories.shared);
        rarefaction.single(shared=all_bacteria.all_categories.shared, outputdir=data/process)"

mothur "#set.dir(input=data/mothur, output=data/mothur);
        make.shared(list=all_bacteria.filter.unique.precluster.an.list, group=bacteria.env_category.groups, label=unique-0.03-0.05-0.10-0.20);
        system(mv data/mothur/all_bacteria.filter.unique.precluster.an.shared data/mothur/all_bacteria.env_category.shared);
        rarefaction.single(shared=all_bacteria.env_category.shared, outputdir=data/process)"

mothur "#rarefaction.single(list=data/mothur/all_bacteria.filter.unique.precluster.an.list, label=unique-0.03-0.05-0.10-0.20, outputdir=data/process);"

rm data/mothur/*rabund data/mothur/mothur*logfile mothur*logfile
```

>>>>

Notes...
* Will want to get rid of sequences with large number of Ns in them - how large?

@@

spread_sheet_good$date <- "2013-07-25"

write.table(spread_sheet_good[,colnames(metadata_good)], file="data/mothur/bacteria.good.metadata", quote=TRUE, na="NA", sep="\t", col.names=F, append=T)

#we'll output the sequence names so the sequences aren't included in the
#rarefaction analysis
write(rownames(spread_sheet_good), file="data/mothur/bacteria.sc_emirge.accnos")
```

@@

write.table(bacteria_emirge_metadata, file="data/mothur/bacteria.good.metadata", quote=TRUE, sep="\t", col.names=F, append=T)

#we'll output the sequence names so the sequences aren't included in the
#rarefaction analysis
write(rownames(bacteria_emirge_metadata), file="data/mothur/bacteria.sc_emirge.accnos", append=T)
```

@@

```{r}
#here we'll read in the list file and extract the OTU assignments for the
# unique, 0.03, 0.05, 0.10, 0.20 cutoffs
list_text <- scan(file="data/mothur/all_bacteria.filter.unique.precluster.an.list", sep="\n", what="")[c(2,5,7,12,22)]
names(list_text) <- c("0.00", "0.03", "0.05", "0.10", "0.20")

#let's remove the otu label and the number of OTUs at each cutoff
parseList <- function(line){

	list_otus <- unlist(strsplit(line, "\t"))[-c(1,2)]
	n_otus <- length(list_otus)

@@

#let's get the metadata file and paste in the OTU and taxonomy data
metadata <- read.table(file="data/mothur/bacteria.good.metadata", header=T, stringsAsFactors=FALSE)
metadata <- cbind(metadata, otu_data[as.character(rownames(metadata)),])
metadata$taxonomy <- taxonomy_data[as.character(rownames(metadata))]

@@

#correct typos:
metadata$category[metadata$category == 'AQD'] <- 'AQO'
metadata$category[metadata$category == 'AQHS'] <- 'AQH'
metadata$category[metadata$category == 'AQS'] <- 'AQMS'
metadata$category[metadata$category == 'VD'] <- 'BD'
metadata$category[metadata$category == 'BM'] <- 'BI'
metadata$category[metadata$category == 'AH'] <- 'AQH'
metadata$category[metadata$category == 'BDBD'] <- 'BD'

#... and we pool the categories by environment
metadata$environment <- rep(NA, nrow(metadata))
metadata$environment[metadata$category == "AE"] <- "aerosol"
metadata$environment[metadata$category == "OT"] <- "other"
metadata$environment[metadata$category %in% c("AQB", "AQBS", "AQF", "AQFS", "AQH", "AQI", "AQM", "AQMS", "AQO", "AQS")] <- "aquatic"
metadata$environment[metadata$category %in% c("BD", "BF", "BI", "BO", "BP")] <- "built"
metadata$environment[metadata$category %in% c("PO", "PR", "PS")] <- "plant"
metadata$environment[metadata$category %in% c("SA", "SD", "SO", "SP")] <- "soil"
metadata$environment[metadata$category %in% c("ZA", "ZN", "ZO", "ZV")] <- "zoonotic"

#let's flag each sequence to indicate how each was generated
source("code/partition_data.R")
metadata$pcr <- is_pcr(metadata)
metadata$cultured <- is_cultured(metadata)
metadata$single_cell <- is_single_cell(metadata)
metadata$emirge_pcr <- is_emirge_pcr(metadata)
metadata$emirge_metag <- is_emirge_metag(metadata)

#... and we output the metadata file...
write.table(metadata, file="data/process/bacteria.v123.metadata", sep="\t")
```

Viola! We'll upload data/process/bacteria.v123.metadata to FigShare and start analyzing the data for the paper.
